---
title: "Tutorial on Pivot Tables and other Excel tasks you can also do in R - Witch Trials"
author: "Kim Fitter"
output: 
  html_document: 
    toc: true
    toc_float: true
---

Which witches and other Excel tasks.....in R. This tutorial assuming you have already installed R and R studio and had a brief introduction to the R basics. Run each line of code yourself or Knit to HTML to walk through this tutorial.

At the end of the exercise we will have an interactive tabular view of the number of witches tried by country.

```{r setup, include=FALSE}
# This is a comment on the setup chunk, we are using echo=TRUE so that the code is included in the R Markdown document
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
```

## Particularly useful packages for this exercise

```{r load packages}
# Load packages
library(tidyverse)
library(rpivotTable)
```

**Tip** Install the package from CRAN the first time you use the package using  `install.packages("<name of package>")` else you will get an Error that there is no package.

**Tip** The latest package version is sometimes available from GitHub so alternatively install from GitHub using the`devtools` package. A package's GitHub README file typically has installation instructions.

## Import Witch Trial data

The dataset used in this tutorial is of 43,000 people tried for witchcraft across
21 European countries over a period of five-and-a-half centuries used for this [economic journal](https://www.researchgate.net/publication/315533575_Witch_Trials). 

First import the data using the `read_csv` function from the [readr](https://cran.r-project.org/web/packages/readr/index.html) R package which is loaded with the tidyverse, to import the raw data to a data frame called `witchdat`.

A data frame, which is a 2-dimensional object where contents can be different data types.

```{r download data}
# Download data from the author's GitHub url
url <- "https://raw.githubusercontent.com/JakeRuss/witch-trials/master/data/trials.csv"
# The read_csv function also produces default messages describing the parsing of columns which is converting the columns into different types.
witchdat <- read_csv(url)
```
  
Let's take a quick look at the `witchdat` data frame. 

In R there are some packages that are included by default; take a look at the packages available in the Global Environment drop down. One such package is called [base](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html) which includes dim function to retrieve the dimensions, the row and column numbers.

```{r dim}
# Retrieve the number of rows and columns
dim(witchdat)
```

Another included package is [utils](https://cran.r-project.org/web/packages/R.utils/index.html) R package which includes the `str` and `structure` functions to view the basic structure. 

```{r structure}
# Use str function to 'Compactly Display the Structure of an Arbitrary R Object'. All R objects store additional attributes to store metadata about an object. In this str function we set the give.attr to FALSE to exclude these attributes.
str(witchdat, give.attr=FALSE)
# Next we will use structure function to view witchdat 
structure(witchdat)
# There are other ways of looking at your data. Try the glimpse function from the tibble package. Take a look at the Help description with ?glimpse to find out more
glimpse(witchdat)
```
  
This is a fairly large data frame!

It looks like there are character variables, double and integer variables.

##  Format "cells" in R

In Excel we can format "cells". Which functions can we use in R?

A tool for data frame data manipulation is the  [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) R package. R packages typically have vignettes which are tutorials and worked examples using the package. There is useful documentation in the [Introduction to dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html)  vignette.

In order to access the variables we can use a `$*`. Alternatively we can use another method that joins objects and functions using a pipe or `%>%`. This can be useful if we would like to use multiple `dplyr` verbs or functions on an object.

We will be use `<-` to assign the formatting changes to `witchdat`.

```{r format}
# From inspection of witchdat this cryptic name gadm.adm0 refers to country. Use dplyr to rename a variable. 
witchdat <- witchdat %>% 
  rename(country=gadm.adm0)
# Perform data type conversion of the deaths variable - use dplyr mutate to change this variable
witchdat <- witchdat %>% 
  mutate(deaths=as.numeric(deaths))
# or we can use dplyr mutate_all to change types of all integer variables (is.integer) to numeric (as.numeric). With mutate_if we don't need the () with is.integer and as.numeric functions as mutate_if will operate on all the columns for which the first test function is TRUE.
witchdat <- witchdat %>% 
  mutate_if(is.integer,as.numeric)
# View the structure of witchdat again to see these formatting changes
str(witchdat, give.attr=FALSE)
```

Try typing 'as.' in the console to see what other options are available to change data types.

For other formatting we can also format the end products (such as plots), using the functions available in those R packages.

##  "Sort" data in R

In Excel we can sort by columns. Which functions can we use in R?

R is different to Excel that we can also view data manipulations without assigning changes to an object. In the sorting examples here we will just view the changes in-line.

```{r arrange}
# Sort by character using arrange from dplyr. We can also pipe other packages functions such as head from the utils package.
witchdat %>% 
  arrange(city) %>% 
  head()
# This sorting also works on numeric variables, sort by year using arrange from dplyr
witchdat %>% 
  arrange(year) %>% 
  head()
# Or we can sort in descending order
witchdat %>% 
  arrange(desc(year)) %>% 
  head()
```

##  Create a "formula" in R

In Excel we can create formulas in cells. Which functions can we use in R?

```{r formula}
# Remember R as a calculator? This is a simple formula
3+2
# Create a new variable in our dataframe called triedbutnotdied as the difference between two variables (formula after the =). Use the mutate function from dplyr
witchdat <- witchdat %>% 
  mutate(triedbutnotdied = tried-deaths)
# In R missing values are typically NA values. A handy base function to know is is.na(). Try typing ?is.na in the console.  Let's use this function in a new formula to test if a variable is populated
witchdat <- witchdat %>% 
  mutate(missingyear = ifelse(is.na(year),"Yup missing","Not missing"))
# View the structure of witchdat again to see these new variables
glimpse(witchdat)
```

##  Filter a "column"

In Excel we can filter by columns. Which functions can we use in R?

```{r filter}
# Filter which witches were tried in decade 1520 by using dplyr filter. Remember to use == for equals in R, whereas = is an assignment in R
witchdat %>% 
  filter(decade==1520) %>% 
  head()
```

##  Calculate on "columns"

In Excel we can add a calculation such as sum or a mean to a column. Which functions can we use in R?

In R Markdown documents we can "highlight" and comment on insights, calculations or data manipulations.

```{r calculation}
# Sum the total deaths. Remember there are NA values so we need to tell this sum function to ignore these with argument na.rm=TRUE. Using sum on the 'whole column' ie variable witchdat$deaths is intuitively the same way of calculating this sum in Excel
sum(witchdat$deaths,na.rm=TRUE)
# Now use dplyr with the select function to select the deaths variable (also the same as deleting the other columns in Excel) then apply the sum on this object
witchdat %>% 
  select(deaths) %>% 
  sum(na.rm = TRUE)
# Now try the dplyr summarise function and sum function then name this sum object as sum_deaths
witchdat %>% 
  summarise(sum_deaths = sum(deaths, na.rm = TRUE))
```

##  Create a "Pivot Table"

In Excel you can create pivot tables. Which functions can we use in R?

A neat R package that creates interactive pivot tables is [rpivotTable](https://cran.r-project.org/web/packages/rpivotTable/).   Have a look at this package's [vignette](https://cran.r-project.org/web/packages/rpivotTable/vignettes/rpivotTableIntroduction.html).

```{r pivot}
# Create an interactive pivot table using the rpivotTable function and package
rpivotTable(witchdat,rows="country",vals="tried",aggregator="Sum")
```
  
  
**Challenge 1** Can you recreate the pivot table as a new R data frame using `dplyr` functions? (Hint ?group_by)

**Challenge 2** On the pivot table widget move the country above the table so the `country` is a pivot table column. Can you extend your challenge 1 answer with `dplyr`  to recreate.

## Save the "spreadsheet"

In Excel you can save your workbooks and spreadsheets. Which functions can we use in R?

Here are some options in R, depending on what output is needed:  

- The `base` R `write.csv` function to write to a csv file  
- The `readr` package in the tidyverse, includes the function `write_csv`, which is faster than write.csv  
- The `xlsx` package `write.xlsx` to write to an Excel workbook  


```{r save}
# Save dataframe with write_csv from readr package
write_csv(witchdat,"witchdat.csv")
```
  
  
  
***  

## Acknowledgments

  
Thanks to Marcia Ferreira @DrMarciaFe, Chris Beltz @BeltzEcology and Ivan Leung @urganmax for reviewing and providing feedback on this tutorial.

