---
title: "Witch Trials Pivot Tables and other Excel tasks you can also do in R"
output: 
  html_document: 
    toc: true
    toc_float: true
---

Which witches and other Excel tasks.....in R. Are there any others that you find useful?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Particularly useful packages for this exercise

```{r install.packages}
library(tidyverse)
#install.packages("rpivotTable")
library(rpivotTable)
```

## Import Witch Trial data

```{r download data}
# Download data from the authors' repo.
url <- "https://raw.githubusercontent.com/JakeRuss/witch-trials/master/data/trials.csv"

# Use the read_csv function to download the raw data to a dataframe called witchdat
witchdat <- read_csv(url)
```

##  Format "cells" in R

In Excel we can format "cells". Which functions can we use in R?

```{r format}
# Change the name of gadm.adm0 variable
# witchdat$country <- witchdat$gadm.adm0
# or we can use dplyr
witchdat <- witchdat %>% 
  rename(country=gadm.adm0)
# Change the class of the deaths variable
witchdat$deaths <- as.numeric(witchdat$deaths)
# or we can use dplyr mutate to change 1 variable
witchdat <- witchdat %>% 
  mutate(deaths=as.numeric(deaths))
# or we can use dplyr mutate_all to change all integer variables
witchdat <- witchdat %>% 
  mutate_if(is.integer,as.numeric)
```

Try typing 'as.' in the console to see what other options are available to change formats.

For other formatting we can format the end products that we create, so that can be function or package dependent.

##  "Sort" data in R

In Excel we can sort by columns. Which functions can we use in R?

```{r arrange}
# Sort by character using arrange from dplyr. We can also pipe base functions such as head.
witchdat %>% 
  arrange(city) %>% 
  head()
# This also works on numeric variables, sort by year using arrange from dplyr
witchdat %>% 
  arrange(year) %>% 
  head()
# Or we can arrange in descending order
witchdat %>% 
  arrange(desc(year)) %>% 
  head()
```

##  Create a "formula" in R

In Excel we can create formulas in cells. Which functions can we use in R?

```{r formula}
# Create a new variable in our dataframe with mutate from dplyr. 
witchdat <- witchdat %>% 
  mutate(triedbutnotdied = tried-deaths)
# In R missing values are typically N/A values. A handy base function to know is is.na(). Try typing ?is.na in the console.  Let's use this function in a new formula to test if a variable is populated
witchdat <- witchdat %>% 
  mutate(missingyear =ifelse(is.na(year),"Yup missing","Not missing"))
```

##  Filter a "column"

In Excel we can filter by columns. Which functions can we use in R?

```{r filter}
# Filter which witches were tried in decade 1520 by using dplyr filter. Remember to use == for equals in R, whereas = is an assignment in R
witchdat %>% 
  filter(decade==1520) %>% 
  head()
```

##  Calculate on "columns"

In Excel we can add a calculation such as sum or a mean to a column. Which functions can we use in R?

```{r calculation}
# Sum the total deaths. Remember there are NA values so we nee dto tell this sum function to ignore these with na.rm=TRUE
sum(witchdat$deaths,na.rm=TRUE)
```

##  Create a "Pivot Table"

In Excel you can create pivot tables. Which functions can we use in R?

R packages typically have vignettes which are tutorials and worked examples using the package. Have a look at this package's [vignette](https://cran.r-project.org/web/packages/rpivotTable/vignettes/rpivotTableIntroduction.html).

```{r pivot}
# Create an interactive pivot table using the rpivotTable function and package
rpivotTable(witchdat,rows="country",cols="deaths",vals="Freq")
```

## Save the "spreadsheet"

In Excel you can save your spreadsheet. Which functions can we use in R?

```{r save}
# Save dataframe with write_csv from readr package
write_csv(witchdat,"witchdat.csv")
```



