---
title: "ggplot2"
author: "Sarah Romanes"
date: "18 June 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
# Download data from the authors' repo.
url <- "https://raw.githubusercontent.com/JakeRuss/witch-trials/master/data/trials.csv"

# Use the base read.csv function to download the raw data to a dataframe called db
db <- read.csv(url)
head(db)
```

What questions can we ask from this dataset?

A few that come to my mind:

* What source provided the most data to this study?
* Is there a relationship between the number of tried/deaths? Did all trials result in deaths?
* Is there a peak decade in which the trials occured? When did it start to ramp down?
* Were some countries more prollific in their trials than others?


# Simple Bar plot (1 qualitative variable)
```{r}
p <- ggplot(db, aes(x=record.source))
p <- p + geom_bar()
p
```

As we can see, the text is super squished on the x axis! We can use `element_text` to fix this.

```{r}
p <- p + theme(axis.text.x = element_text(angle=90, hjust = 1))
p
```


# Histogram (1 quantitative variable)

```{r}
p1 <- ggplot(db, aes(x=decade))
p1 <- p1 +geom_histogram()
p1
```


We can spice it up by adding an aesthetic (qualitative variable)

```{r}
p1 <- p1 +geom_histogram(aes(fill=gadm.adm0))
p1
```

# Simple Scatterplot [2 quant]

```{r}
p2 <- ggplot(db, aes(x=tried, y=deaths))
p2 <- p2+geom_point()
p2
```


# Boxplot [1 quant, divided by 1 qual]

```{r}
p3 <- ggplot(db, aes(x=gadm.adm0, y=deaths))
p3 <- p3+geom_boxplot() + theme(axis.text.x = element_text(angle=90, hjust = 1))
p3
```

# Boxplot [1 quant, divided by 1 qual] with scaled y axis

```{r}
p3 + scale_y_log10()
```

